Epistasis is widespread in the genetic control of transcription in humans

### 2013-06-07 docs/manuscript/nature_letter.tex
LD check - as expected in BSGS therefore no evidence for population stratification

\begin{abstract}
A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Carlborg2004, Hill2008a, Crow2010}. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design where any effects are expected to be large \cite{Powell2012}, evidence for multiple instances of epistasis is found. In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 434 significant pairwise epistatic interactions acting on the expression levels of 165 genes. Twenty-nine interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing. Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects ($\lambda_{GC} = 1.35$, $p = 0.04$). We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of known regions of \emph{in vitro} chromosome interactions \cite{LiebermanAiden2009} ($p < 1.8 \times 10^{-10}$). Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs. For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first strong evidence for the widespread existence of epistatic genetic effects emerging from natural genetic variation in humans, and makes inference on its contribution.
\end{abstract}

@@

% \subsection{Introduction}
The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits. Typically, SNPs are modelled as contributing linearly, independently, and cumulatively to the mean of a trait and this has been successful in identifying thousands of causal variants \cite{Visscher2012}. To date there is little convincing empirical evidence for epistasis playing a substantial role in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its importance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}. Yet outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}. Here we demonstrate that instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.

Detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009}. When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two ways to overcome this problem, either by using extremely large sample sizes \cite{LangoAllen2010}, or by analysing traits that are likely to have large effect sizes. Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that any genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.

% \subsection{Initial search in discovery set}
In a dataset of 846 individuals genotyped at 528509 SNPs, we tested every pair of SNPs for genetic interactions on each of 7339 expression traits in whole blood. After stringent filtering and multiple testing correction we had identified 501 putative genetic interactions influencing 238 gene expression levels. Of these, 434 passed filtering in two independent datasets of which 29 had interaction terms surpassing a Bonferroni corrected level of significance (Figure \ref{fig:gpmaps}, Table \ref{tab:bonferroni}).

% \subsection{Replication}
Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average four orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$). We observed substantial replication of interaction signals in both datasets. Of the 480 original pairs that passed filtering (Methods) in Fehrmann and EGCUT, 421 of the SNP pairs showed replication for the interaction test at the 5\% false discovery rate (FDR) level in at least one dataset, and 148 in both datasets (Supplementary Figure \ref{fig:qqplotfdr}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations. We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and 18 were significant in both. The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}). A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication.

@@

\subsection{Methods Summary}
We searched for pairwise epistasis exhaustively in the Brisbane Systems Genetics Study (BSGS) dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes. Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}. Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone. All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs. Of these, 434 passed filtering in both replication datasets.

>>>>

\begin{abstract}
A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Carlborg2004, Hill2008a, Crow2010}. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, evidence for multiple instances of epistasis is found. In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 223 genes. Twenty-nine interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing. Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$). We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{LiebermanAiden2009} ($p < 1.8 \times 10^{-10}$). Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs. For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first strong evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans, and demonstrates its significant contribution to the variance of gene expression.
\end{abstract}

@@

% \subsection{Introduction}
The past decade has seen a tremendous amount of activity in mapping genetic polymorphisms that underlie complex traits. Typically, SNPs are treated using an additive model, assumed to contribute linearly, independently, and cumulatively to the mean of a trait. This has been successful in identifying thousands of causal variants \cite{Visscher2012}, but to date there is little empirical exploration of the role that epistasis plays in the architecture of complex traits in humans \cite{Strange2010, Evans2011}, though its importance is frequently the subject of debate \cite{Carlborg2004, Hill2008a, Crow2010}. Outside the prism of association studies there is widespread evidence for epistasis, not only at the molecular scale from artificially induced mutations \cite{Costanzo2010} but also at the evolutionary scale in fitness adaptation \cite{Weinreich2006} and speciation \cite{Breen2012}. Here we demonstrate that multiple instances of statistical epistasis can be detected in human traits, and that some quantification of its relative contribution to phenotypic variance can be made.

Detection of epistasis is hampered by power issues for several reasons, including increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009}. When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two ways to overcome this problem, either by using extremely large sample sizes \cite{LangoAllen2010}, or by analysing traits that are likely to have large effect sizes. Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that many genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In a dataset of 846 individuals genotyped at 528509 SNPs, we tested every pair of SNPs for genetic interactions on each of 7339 expression traits in whole blood. After stringent filtering and multiple testing correction we had identified 501 putative genetic interactions influencing 238 gene expression levels. Of the 501 discovery interactions, data was available for 434 in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 29 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}), demonstrating that these complex genetic patterns are common even across independent populations. Additionally, there was extreme enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication. We observed that 316 of the remaining 405 discovery SNPs had replication interaction $p$-values exceeding the 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$). A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication. Though a necessary step to establish the veracity of the signals from the discovery set, it is theoretically difficult to replicate epistasis because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. Therefore

We also report that 39 of the interactions from the discovery set were significant at the Bonferroni level in at least one replication set (Supplementary Figure \ref{fig:qqplotbonf}), and 18 were significant in both. The GP maps for these interactions are remarkably similar in all three datasets (Figure \ref{fig:gpmaps}).

@@

\subsection{Methods Summary}
We searched for pairwise epistasis exhaustively in the Brisbane Systems Genetics Study (BSGS) dataset \cite{Powell2012}, which comprises 846 individuals who are genotyped at 528509 autosomal SNPs and who have gene expression levels measured in whole blood samples for 7339 probes representing 6158 RefSeq genes. Recent hardware and software \cite{Hemani2011} advances made it possible to perform the $1.03 \times 10^{15}$ statistical tests to complete this analysis. We used permutation analysis \cite{Churchill1994a} to calculate an experiment-wide significance threshold of $2.91 \times 10^{-16}$ at the 5\% family-wise error rate (FWER). SNP pairs were modelled for full genetic effects, including marginal additive and dominance at both SNPs plus four interaction terms. Though we could have used a less complex model to improve statistical efficiency, we deemed it important to be agnostic about the type of epistasis that might exist, and therefore chose not to over-parameterise the test \cite{Marchini2005, Hemani2013}. Because there are many large marginal effects present in these data it was necessary to perform several filtering steps to exclude SNP pairs that were significant due to marginal effects alone. All SNP pairs with LD $r^2 > 0.1$ were removed, and were required to have at least five data points in all nine genotype classes. If multiple SNP pairs were present on the same chromosomes for a particular expression trait then only the sentinal SNP pair was retained. Finally, a nested test contrasting the full genetic model against the marginal additive and dominance model was performed for each remaining SNP pair (Methods), resulting in 501 significant interactions after Bonferroni correction for multiple testing of the filtered SNPs. The significant SNP pairs were carried forward for replication in two independent datasets that used the same expression assays for analysing transcription in whole blood, the Fehrmann dataset \cite{Fehrmann2011} ($n=1240$) and the Estonian Genome Centre University of the University of Tartu (EGCUT) dataset \cite{Metspalu2004} ($n=891$). Of these, 434 passed filtering in both replication datasets.
